Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan

### 2024-08-02 analysis/paper/paper.qmd
text editting

# Introduction
<!-- introduction about European expansion -->
European expansion since the 15th century marked the beginning of an era characterized by global trade, exploration, and colonization, which often led to profound impacts on indigenous societies in many parts of the world. The forms of European colonization and power relations varied across regions, depending on specific spatial and temporal contexts [@Jordan2009]. Understanding the dynamics of these colonial processes has become a significant issue in archaeology, as it provides critical insights into the complex outcomes of European colonization [@Gosden2004]. In Asia, European colonization was primarily driven by the procurement of Asian luxury goods, raw materials, or natural resources based on commercial interest [@Prakash1997]. The new exchange systems and governance policies imposed by European colonizers often had profound effects on local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002]. However, Indigenous communities did not just passively accept European colonization. Recent discussions on colonial interactions also highlight the diverse interactions employed by Indigenous communities in response to European presence, such as resistance, negotiation, persistence, resilience, and entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jordan2009; @Alexander1998].

<!-- NE Taiwan as a case study -->
While many studies focus on prominent colonies, places on the periphery of colonial center were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017]. Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion. Northeastern Taiwan, specifically Yilan, serves as an example of a peripheral case in the 17th century when Europeans established their colonial base in northern Taiwan for a better access to Asian market. Due to its geologically challenging and relatively inaccessible location, Yilan experienced limited direct European control. However, local settlements were frequently involved in the regional exchange network stimulated by the colonial presence in the north. In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export in a large scale, involving many Indigenous communities and Han Chinese settlers [@Heyns2001; @Cheng2017]. Despite some discussions on the policies and operations of the deerskin trade, direct investigations on archaeological evidence into the experiences of local indigenous people to the commercial activities are rare. The European impacts on local ecological conditions have been recorded in many places in colonial settings [@Jorden2009]. Whether the interaction between Indigenous people and their environment changed due to European trade demands requires further archaeological investigation.

<!-- research aim and questions -->
In this paper, I will explore the interactions between people and their environment in a colonial context by examining resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), a largest settlement in northeastern Taiwan [@Chen2007]. Kiwulan experienced periods of European presence in the 17th century, followed by a large influx of Han Chinese migration in the 19th century. I will primarily focus on subsistence practices, as the choices made in these practices can reveal the strategies Indigenous people employed to cope with European colonial activities and the associated changes in their environment. My research questions are: What were the subsistence practices of Indigenous people before and after the European presence? Can we identify shifts in these strategies in response to colonial economic demands? To address these questions, I will examine the faunal and plant remains from Kiwulan, alongside a preliminary analysis of ceramic residues for investigating culinary practices. Through an environmental perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.

# European encounters in the northern region of Taiwan
<!-- the Spanish colonization and the power relations in particular spatial and temporal contexts-->
In 1626, the Spanish East Indies Company (Indias Orientales Españolas) arrived in northern Taiwan and established the San Salvador colony at Heping Dao in Keelung, followed by Santo Domingo in Tamsui, to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007]. Despite the intention to establish a colonial city, San Salvador remained in its early developmental stages, primarily consisting of a fortress and a church, with a population of Spanish officers, soldiers, and missionaries [@Borao2009, pp. 103-134; @Berrocal2018]. The area neighboring San Salvador was inhabited by the Taparris and Quimaurris, local indigenous groups who had resided at the location of the fort before being expelled by the Spanish [@Borao2001, SIT:165, p.81]. During Spanish colonization (1626-1642), the colonial processes may not have involved extensive colonialism or asymmetrical power dynamics, with territorial control likely limited to the fort. Written records mention that the Spanish appeared to rely on the local indigenous people for exchanging local goods and food supplies [@Borao2001, SIT:170-171]. This is further supported by archaeological evidence at San Salvador [@Berrocal2018], where only a few European materials were found, alongside more indigenous and Chinese items that might have been used by the Europeans, indicating cultural entanglement.

<!-- the Dutch domination and the processes by which it was established, how and why colonial powers were able to exert control -->
In 1642, the Dutch defeated the Spanish and took over the San Salvador and the Santo Domingo, expanding their colonial power to northern and eastern Taiwan until 1662. Compared to the Spanish, the Dutch were more strategic in consolidating their power by forming alliances with indigenous villages, implementing political and economic institutions, and conquering villages that were unwilling to comply [@Kang2016; @Andrade2007]. For example, the VOC governors used various colonial institutions to secure their authority, including annual landdag gathering, taxation, tribute system, village lease system, census investigations, and the selection of indigenous representatives [@Kang2016; @Andrade1997; @Andrade2007]. Among the political institutions, the tribute system operated from 1642 to 1647 and village lease system introduced in 1644 had great impacts on local indigenous economy and environment. Based on tribute system, each household of allied villages was required to pay rice or high-quality deerskin annually. The amount of deerskins collected from indigenous people in northern Taiwan once reached 7,000 pieces in 1646 [@Andrade2007]. Moreover, to increase deerskin exports to Japan, the Dutch implemented a hunting license system (1636-1645) that gave Han Chinese permission to hunt deer on indigenous lands during certain periods [@Heyns2001]. From 1633 to 1661, the average of deerskins exported from Taiwan was about 67,000 per year [@Koo2011; @Cheng2017]. Those colonial activities accelerated the decline in the deer population and could lead to a harsh situation for the indigenous economy .

<!--the variation of colonial powers over native populations, indirect colonization at Kavalan-->
Although Dutch power expanded across nearly all of Taiwan in the mid-17th century, their colonial control and domination varied significantly in different geographical regions. In areas distant from the European colonial bases, colonial control was often challenging and largely unsuccessful. Yilan in northeastern Taiwan is an example of minimal European presence and indirect colonization, with fewer imposed colonial rules for most of the time [@Kang2012]. Yilan is an alluvial plain located about 70 kilometers from the European colony in Keelung, separated by mountains, and inhabited mainly by the local indigenous group, the Kavalan. Given its strategic location on the route to Torrobouan, where the Dutch believed gold mines existed, they were eager to bring this area under their control [@DZI2000; @Kang2016, p47]. In 1644, the Dutch invaded Yilan to incorporate the defeated Kavalan villages into their northern colonial territory, and demanded tribute from the indigenous people [@Kang2016]. nstead of engaging directly with the Kavalan people, the Europeans relied heavily on the assistance of the indigenous Basay people from Keelung to coordinate and collect tributes in Yilan [@Kang2016]. This indirect approach allowed the maintenance of autonomy and flexibility of Kavalan people for negotiation tributes, such as deerskins or products, requested by the VOC [@Kang2016].

<!-- pre-exiting trade network -->
Before the arrival of the Europeans in northern Taiwan, an inter-insular trade network had been established since the early 15th century, involving Chinese merchants and indigenous groups in Tamsui and Keelung [@Chen2005]. Chinese traders transported goods such as textiles, porcelains, beads, and metal items to exchange for raw materials from Taiwan, such as gold, sulfur, and animal hides and meats from the local Basay people [@Chen2005; @Andrade2007]. This regional trade network was further connected to maritime trade between Ming China and the Ryukyu Kingdom during the 15th and 16th centuries and later to European long-distance trade in the 17th century [@Chen2005; @Von2019; @Wang2007].Yilan was included in this long-established trade network, where the Kavalan people exchanged products such as rice and deerskins for foreign goods and handicrafts from the Basay people [@Kang2003]. This is evidenced by the large amount of imported foreign goods found across archaeological sites in Yilan, indicating frequent trade activities, especially around the 17th century [@TaiwanHistory2005]. Yilan experienced frequent trading activities driven by the pursuit of foreign goods even before the Europeans arrival, and these activities continued during the European presence in northern Taiwan, likely with some occasional pressure from European powers. This unqiue historical and spatial contexts make Yilan an ideal location to study indigenous reactions to foreign influences and the associated trade networks from a long-term perspective in a pericolonial context [@Acabado2017; @Kang2016, pp. 31-78; @Trabert2017].

# Kiwulan site and previous studies
This study focuses on archaeological evidence from Kiwulan, a large multi-component site of with an upper component spanning from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007]. Kiwulan was inhabited by Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture and archaeological artifacts [@Chen2004]. According to a mid-17th-century Dutch census, there were about 40 indigenous villages with approximately 10,000 people, and Kiwulan (referred to as Kibannoran in Dutch records) was the largest village, with 160 households and 711 people [@Nakamura1938]. In the early 19th century, the population of Kiwulan dramatically decreased to fewer than 100 people due to the population pressure of Han Chinese migrants, which led to the relocation of Kavalan people to eastern Taiwan [@Zhan1995].

The excavation at Kiwulan uncovered 262 units (4 x 4 m) and revealed abundant artifacts and features, such as middens, post-holes, and burials. These findings suggest a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004]. In addition to locally made pottery, porcelains, stonewares, glass beads, carnelian beads and metal artifacts were largely found in the residential area and burials. Since there is no evidence of local production of beads or metallurgy, these goods are believed to have been acquired through a regional exchange network with people from outside the area.

# Previous research of

>>>>

# Introduction
<!-- introduction about European expansion -->
European expansion since the 15th century was usually associated with long-distance trade, exploration, and colonization in many parts of the world, which often led to profound impacts on indigenous societies. Depending on particular geographic and historical circumstances, many regions experienced different forms of European colonialism and power relations [@Jordan2009]. Understanding the dynamics of these colonial processes has become a significant issue in archaeology because it provides critical insights into the complex outcomes of European colonization [@Gosden2004]. In Asia, European colonization was mainly motivated by the acquisition of raw materials, natural resources, or luxury items from Asia for commercial purposes [@Prakash1997]. Although the new exchange systems or colonial institutions introduced by Europeans could change local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002], indigenous groups did not just accept European colonization passively. Recent discussions on colonial interactions also suggest the active role of indigenous groups in responding to the arrival of Europeans and presenting resistance, negotiation, persistence, resilience, and entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jordan2009; @Alexander1998].

<!-- NE Taiwan as a case study -->
While many studies focus on prominent colonies, places on the periphery of colonial centers were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017]. Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion. Yilan, located in northeastern Taiwan, is an example of a periphery region where European control was limited and indirect for most of the time but intensively involved in regional trade activities stimulated by the European presence. In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export on a large scale, involving many Indigenous communities and Han Chinese settlers [@Heyns2001; @Cheng2017]. Despite some discussions on the operations and policies of the deerskin trade, direct investigations on archaeological evidence into the experiences of local indigenous people with the commercial activities are rare. The European impacts on local ecological conditions have been recorded in many places in colonial settings [@Jorden2009]. Whether the interaction between Indigenous people and their environment changed due to European trade demands requires further archaeological investigation.

<!-- research aim and questions -->
In this paper, I will explore the interactions between people and their environment in a colonial context by examining the resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), the largest settlement in northeastern Taiwan [@Chen2007]. Kiwulan encountered European presence in the 17th century, followed by a large influx of Han Chinese migration in the 19th century. What was the resource use of Indigenous people before and after the foreign presence? Can we identify shifts in these practices in response to colonial economic demands? To address these questions, I will focus on subsistence practices because the choices made in the practices can reveal the strategies Indigenous people used to respond to European exploitation with associated changes in their environment. I will primarily examine the zooarchaeological evidence from Kiwulan for subsistence practices, coupled with ceramic residues for culinary practices. Through an environmental perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.

# European encounters in the northern region of Taiwan
<!-- the Spanish colonization and the power relations in particular spatial and temporal contexts-->
In 1626, the Spanish East Indies Company (Indias Orientales Españolas) arrived in northern Taiwan and established the San Salvador colony at Heping Dao in Keelung, followed by Santo Domingo in Tamsui, to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007]. Although the Spanish intended to build a colonial city, San Salvador remains its early development during their colonization, primarily consisting of a fortress and a church [@Borao2009, pp. 103-134; @Berrocal2018]. The neighboring area of the fort was inhavited by local indigenous groups, the Taparris and the Quimaurris, who had lived at the location of the fort before being expelled by the Spanish [@Borao2001, SIT:165, p.81]. The relations between the Spanish and indigenous people might not involve extensive colonialism or intercultural domination, and the Spanish territorial control could be limited to the fort. Instead, the Spanish relied on the exchange of local goods and food surpluses with local indigenous people, according to written records [@Borao2001, SIT:170-171]. This corresponds with a higher proportion of indigenous and Chinese items with little European material found in San Salvador [@Berrocal2018], indicating cultural entanglement.

<!-- the Dutch domination and the processes by which it was established, how and why colonial powers were able to exert control -->
In 1642, the Dutch defeated the Spanish and took over San Salvador and Santo Domingo, expanding their colonial power to northern and eastern Taiwan until 1662. Compared to the Spanish, the Dutch were more strategic in consolidating their power by forming alliances with indigenous villages, implementing political and economic institutions, and conquering villages that were unwilling to comply [@Kang2016; @Andrade2007]. For example, the VOC governors used various colonial institutions to secure their authority, including the annual landdag gathering, taxation, tribute system, village lease system, census investigations, and the selection of indigenous representatives [@Kang2016; @Andrade1997; @Andrade2007]. Among the political institutions, the tribute system operated from 1642 to 1647 and the village lease system introduced in 1644 had great impacts on the local indigenous economy and environment. Based on the tribute system, each household in allied villages was required to pay rice or high-quality deerskin annually. The amount of deerskins collected from indigenous people in northern Taiwan once reached 7,000 pieces in 1646 [@Andrade2007]. Moreover, to increase deerskin exports to Japan, the Dutch implemented a hunting license system (1636-1645) that gave Han Chinese permission to hunt deer on indigenous lands during certain periods [@Heyns2001]. From 1633 to 1661, the average number of deerskins exported from Taiwan was about 67,000 per year [@Koo2011; @Cheng2017]. Those colonial activities accelerated the decline in the deer population and could lead to a harsh situation for the indigenous economy.

<!--the variation of colonial powers over native populations, indirect colonization at Kavalan-->
Although the Dutch power expanded across nearly all of Taiwan in the middle of the 17th century, the extent of their colonial dominance and control differed greatly among different regions. Colonial rules in regions far from the European forts were usually difficult and unsuccessful. Yilan in northeastern Taiwan is an example of minimal European presence and indirect colonization, with fewer imposed colonial rules for most of the time [@Kang2012]. Yilan is an alluvial plain, located about 70 kilometers from the European colony in Keelung, with mountains in between. The main inhabitants in Yilan were the local indigenous group, the Kavalan. Given its strategic location on the route to Torrobouan, where the Dutch believed gold mines existed, they were eager to bring this area under their control [@DZI2000; @Kang2016, p47]. In 1644, the Dutch invaded Yilan to incorporate the defeated Kavalan villages into their northern colonial territory and demanded tribute from the indigenous people [@Kang2016]. Instead of engaging directly with the Kavalan people, the Europeans relied heavily on the assistance of the Basay people active in northern Taiwan to coordinate and collect tributes in Yilan [@Kang2016]. This indirect approach allowed the maintenance of autonomy and flexibility for the Kavalan people, who were able to negotiate tribute requested by the VOC [@Kang2016].

<!-- pre-exiting trade network -->
Before the arrival of the Europeans in northern Taiwan, an inter-insular trade network had been established since the early 15th century, involving Chinese merchants and indigenous groups in Tamsui and Keelung [@Chen2005]. Chinese traders transported goods such as textiles, porcelains, beads, and metal items to exchange for raw materials from Taiwan, such as gold, sulfur, and animal hides and meats from the local Basay people [@Chen2005; @Andrade2007]. This regional trade network was further connected to maritime trade between Ming China and the Ryukyu Kingdom during the 15th and 16th centuries and later to European long-distance trade in the 17th century [@Chen2005; @Von2019; @Wang2007].Yilan was included in this long-established trade network, where the Kavalan people exchanged products such as rice and deerskins for foreign goods and handicrafts from the Basay people [@Kang2003]. This is evidenced by the large amount of imported foreign goods found across archaeological sites in Yilan, indicating frequent trade activities, especially around the 17th century [@TaiwanHistory2005]. Yilan experienced frequent trading activities driven by the pursuit of foreign goods even before the Europeans arrival, and these activities continued during the European presence in northern Taiwan, likely with some occasional pressure from European powers. This unique historical and spatial context makes Yilan an ideal location to study indigenous reactions to foreign influences and the associated trade networks from a long-term perspective in a pericolonial context [@Acabado2017; @Kang2016, pp. 31-78; @Trabert2017].

# Archaeological context at Kiwulan
Kiwulan, located in the northern part of Yilan, is a multi-component site, where the upper component spans from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007]. As a salvage project, a large scale excavation was conducted at Kiwulan to recover information before it was destroyed. A total of 262 excavation units uncovered abundant artifacts and features, including middens, burials, and post-holes with wooden posts in situ. It suggests that Kiwulan is a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004]. The main inhabitants of Kiwulan were the Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture recorded in historical records and archaeological artifacts [@Chen2004]. According to a mid-17th-century Dutch census, there were about 40 indigenous villages with approximately 10,000 people, and Kiwulan (referred to as Kibannoran in Dutch records) was the largest village, with 160 households and 711 people [@Nakamura1938]. In the early 19th century, the population of Kiwulan dramatically decreased to fewer than 100 people due to the population pressure of Han Chinese migrants, which led to the relocation of Kavalan people to eastern Taiwan [@Zhan1995].

The most common artifact was locally made pottery with geometric patterns as surface treatments, which was also found at many simultaneous sites in Yilan. In addition to pottery, a variety of foreign goods, such as porcelains, stonewares, glass beads, carnelian beads, and metal artifacts, were largely found in both the residential area and burials. Due to no evidence of local production of beads or metallurgy, these items are believed to have been acquired through a regional exchange network, primarily with the Basay people active in northern Taiwan. This indicates that Kiwulan has continuously participated in a regional trade network closely associated with northern Taiwan since the 15th century and influenced by the long-distance trade during the 17th century [@]. Moreover, the use of foreign goods appears to have some cultural significance, and previous studies suggested a possible increase in social differentiation due to the introduction of foreign goods [@Wang2020ornament; @Wang2021].

There were a wide range of animal bones and plant remains uncovered, reflecting the diversity of the local diet and the environment close to river and coastal area. The animals include deer, pigs, cows, birds, fish, shellfish, and plants mainly consisting of rice, peach, pandanus, and cucurbits [@Chen2004]. Rice farming seems to be common but limited to a household scale, according to historical records [@]. Shellfish consist of bivalves such as clams and oysters and univalves from rivers and brackish water environments, which indicates a reliance on aquatic and marine resources. Also, those archaeological remains correspond with tools related to subsistence found at Kiwulan, including iron arrowheads for hunting, iron sickles for crop harvesting, and wooden paddles for water-related activities. According to those pieces of evidence, the subsistence practices at Kiwulan should be primarily reliant on rice farming, supplemented by hunting and fishing for wild resources.
